Interpretable Neural Networks using EAGGA

### 2025-03-15 paper/main.tex
remove todos

\section{Introduction}
Regularisation has been extensively shown to improve performance from linear models to neural networks.
One of the most popular types of regularisation, Ridge or L2-regularisation is still the preferred method for handling
multicollinear data in linear models \citep[p. 55]{ridge}.
Practicioners training neural networks regularly use regularisation methods such as dropout to reduce
co-adaptation \citep[p. 1]{hinton2012improvingneuralnetworkspreventing} or early stopping to avoid overfitting on the training data \citep[p. 778f]{early_stopping}.
\\
Some regularisation can even increase interpretability, such as LASSO, or L1-regularisation for linear models, which reduces the number of
features used to fit the model by setting some coefficients to 0 \citep[p. 267]{lasso}.
\\
Neural networks as a model class have achieved impressive performance on a multitude of problems, yet still struggle with the most common format --
tabular data \citep[p. 7499]{Borisov_2024}.
Recent research, however, suggest that strong regularisation is beneficial to a neural network's performance on tabular data \citep[8]{NEURIPS2021_c902b497}.
\citet{EAGGA} propose a framework to increase the interpretability of XGBoost using regularisation.
In this report, we explore if the same framework can be extended to neural networks to tackle both interpretability and achieve performance on tabular data
that is on par with XGBoost.

@@

% Taxonomy
\citet[chap. 2]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs),
a model class that in its fully-connected feedforwad form is inherently non-interpretable. % TODO: letztes halbsatz vllt lieber in intro unterbringen, hier ist der awkward und unnötig ausschmückend
\\
\textbf{Passive vs Active Approaches}, where \textit{passive} are all post-hoc methods and \textit{active} methods actively change either the architecture or
training process to increase model interpretability.
\\
\textbf{Type of Explanations}, distinguishing between \textit{example} methods, providing concrete examples of what leads to a desired output, \textit{attribution}
methods that attribute the effect on the output for a specific feature, \textit{hidden semantic} methods, which explain the types of inputs particular neurons or layers
pick up on, and logical \textit{rules}, such as if-then clauses or tree-induced rules.
\\
\textbf{Local vs Global Interpretability}, ranging from \textit{local} methods providing explanations based on individual samples, \textit{semi-local} methods,
explaining model behaviour for sets of samples grouped by some criterion, to \textit{global} methods, which explain the network as a whole.

@@

\begin{equation}  % TODO: take out

@@

\begin{equation}  % TODO: make inline or provide further information

@@

\subsection{Neural Architecture Search}
A special case of HPO is Neural Architecture Search (NAS). Because of the flexibility of neural network architectures (number of hidden layers, strength of dropout,
each layer can have a different numer of neurons, each layer can have a different activation function, etc.), traditional HPO methods cannot efficiently explore the
entire space.
In our extension of the EAGGA approach we don't employ NAS as research focusses on the NLP and image domains, % TODO: find some more references
where features (i.e. tokens or pixels, respectively) exhibit strong correlation amongst themselves, which is usually not the case for tabular data \citep[p. 7499]{Borisov_2024}.
Still, we want to outline two notable approaches for neural architecture search in this paper, one of which are \textbf{cell search spcaces}.
These modularise an NN into a chain-structure of cells, where each cell is a basic building block of the respective architecture. For regular feedforwad neural networks,
this could be a linear layer of fixed size and with a specific activation function,
for CNNs this could be a specific convolutional operation or pooling function, for instance.
Instead of tuning every hyperparameter, the sequential placement of the cells is then optimzed during NAS, e.g. via random search or Bayesian
Optimization. \citep[chap. 3.2]{elsken_neural_2019}
\\
Treating a neural network as a sequence of cells instead of one fixed entity further allows exploiting this structure with special sequential optimization techniques.
\citet[p. 3]{zoph2017neuralarchitecturesearchreinforcement} and \citet[pp. 2-4]{Zoph_2018_CVPR} propose a cell search method using RNNs and reinforcement learning.
The cells in their approach are not fixed, but have hyperparameters themselves, such as the filter size and the stride size of a convolutional layer.
These hyperparameters are then predicted sequentially, given all the already predicted hyperparameters of previous layers. The RNN is trained using reinforcement
learning, where the actions are the different values the network can predict for a given hyperparameter and the reward function is the RNN's performance on held-out
validation data.
\\
The second approach to NAS is the \textbf{one-shot model}, which trains a ``fabric'', which can be understood as a DAG comprising all architectures of a given
search space. Each path through the DAG is one specific network architecture, where the nodes represent a layer of neurons and the edges in-between are operations
on the layers' values. The entire DAG has one designated input and one output node.
It is then trained just like a regular NN, after which the optimal path, i.e. network architecture, is selected from the fabric.
This has the advantage of, albeit being slower than training a single model, being far more efficient and less expensive than training all architectures encoded
in the fabric. \citep[pp. 1-2, p.8]{saxena2017convolutionalneuralfabrics}
% don't necessarily talk about co-adaptation problem

@@

  \caption{TODO: caption}

@@

  \caption{TODO: caption}

@@

\appendix
\section{Software used}
for implementation we used
openml \cite{OpenML}, \cite{OpenMLPython},
numpy \cite{numpy},
pandas \cite{pandas1}, \cite{pandas2},
pytorch \cite{PyTorch},
scikit-learn \cite{scikit-learn},
scipy \cite{SciPy},
pymoo \cite{pymoo}, and
tqdm \cite{tqdm}

>>>>

\section{Introduction}
Regularisation has been extensively shown to improve performance from linear models to neural networks.
One of the most popular types of regularisation, Ridge or L2-regularisation is still the preferred method for handling
multicollinear data in linear models \citep[p. 55]{ridge}.
Practicioners training neural networks regularly use regularisation methods such as dropout to reduce
co-adaptation \citep[p. 1]{hinton2012improvingneuralnetworkspreventing} or early stopping to avoid overfitting on the training data \citep[p. 778f]{early_stopping}.
\\
Some regularisation can even increase interpretability, such as LASSO, or L1-regularisation for linear models, which reduces the number of
features used to fit the model by setting some coefficients to 0 \citep[p. 267]{lasso}.
\\
Neural networks as a model class have achieved impressive performance on a multitude of problems, yet still struggle with the most common format --
tabular data \citep[p. 7499]{Borisov_2024}.
Recent research, however, suggest that strong regularisation is beneficial to a neural network's performance on tabular data \citep[8]{NEURIPS2021_c902b497}.
\citet{EAGGA} propose a framework to increase the interpretability of XGBoost using regularisation.
In this report, we explore if the same framework can be extended to neural networks, a model class that in its fully-connected feedforwad form is
inherently non-interpretable, to tackle both explainability and achieve performance on tabular data that is on par with XGBoost.

@@

% Taxonomy
\citet[chap. 2]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs).
\\
\textbf{Passive vs Active Approaches}, where \textit{passive} are all post-hoc methods and \textit{active} methods actively change either the architecture or
training process to increase model interpretability.
\\
\textbf{Type of Explanations}, distinguishing between \textit{example} methods, providing concrete examples of what leads to a desired output, \textit{attribution}
methods that attribute the effect on the output for a specific feature, \textit{hidden semantic} methods, which explain the types of inputs particular neurons or layers
pick up on, and logical \textit{rules}, such as if-then clauses or tree-induced rules.
\\
\textbf{Local vs Global Interpretability}, ranging from \textit{local} methods providing explanations based on individual samples, \textit{semi-local} methods,
explaining model behaviour for sets of samples grouped by some criterion, to \textit{global} methods, which explain the network as a whole.

@@

@@

@@

\subsection{Neural Architecture Search}
A special case of HPO is Neural Architecture Search (NAS). Because of the flexibility of neural network architectures (number of hidden layers, strength of dropout,
each layer can have a different numer of neurons, each layer can have a different activation function, etc.), traditional HPO methods cannot efficiently explore the
entire space.
In our extension of the EAGGA approach we don't employ NAS as research focusses on the NLP and
image domains \citep{baker2017designingneuralnetworkarchitectures},  \citep{10.5555/3504035.3504375},  \citep{10.1145/3071178.3071229},
where features (i.e. tokens or pixels, respectively) exhibit strong correlation amongst themselves, which is usually not the case for tabular data \citep[p. 7499]{Borisov_2024}.
Still, we want to outline two notable approaches for neural architecture search in this paper, one of which are \textbf{cell search spcaces}.
These modularise an NN into a chain-structure of cells, where each cell is a basic building block of the respective architecture. For regular feedforwad neural networks,
this could be a linear layer of fixed size and with a specific activation function,
for CNNs this could be a specific convolutional operation or pooling function, for instance.
Instead of tuning every hyperparameter, the sequential placement of the cells is then optimzed during NAS, e.g. via random search or Bayesian
Optimization. \citep[chap. 3.2]{elsken_neural_2019}
\\
Treating a neural network as a sequence of cells instead of one fixed entity further allows exploiting this structure with special sequential optimization techniques.
\citet[p. 3]{zoph2017neuralarchitecturesearchreinforcement} and \citet[pp. 2-4]{Zoph_2018_CVPR} propose a cell search method using RNNs and reinforcement learning.
The cells in their approach are not fixed, but have hyperparameters themselves, such as the filter size and the stride size of a convolutional layer.
These hyperparameters are then predicted sequentially, given all the already predicted hyperparameters of previous layers. The RNN is trained using reinforcement
learning, where the actions are the different values the network can predict for a given hyperparameter and the reward function is the RNN's performance on held-out
validation data.
\\
The second approach to NAS is the \textbf{one-shot model}, which trains a ``fabric'', which can be understood as a DAG comprising all architectures of a given
search space. Each path through the DAG is one specific network architecture, where the nodes represent a layer of neurons and the edges in-between are operations
on the layers' values. The entire DAG has one designated input and one output node.
It is then trained just like a regular NN, after which the optimal path, i.e. network architecture, is selected from the fabric.
This has the advantage of, albeit being slower than training a single model, being far more efficient and less expensive than training all architectures encoded
in the fabric. \citep[pp. 1-2, p.8]{saxena2017convolutionalneuralfabrics}
% don't necessarily talk about co-adaptation problem

@@

  \caption{The EAGGA network architecture for a dataset with features A, B, C, D, E, F, G. The group structure might look as follows: $(\{D,G\}, (\{A,B,C\}, 1), (\{E,F\}, 0))$}

@@

  \caption{Visualisation of the data set split. Outer holdout split (2/3), mid 5-fold CV split, inner holdout split (80\%)}

@@

\appendix
\section{Software used}
For the Python implementation of EAGGA we use OpenML \citep{OpenML}, \citep{OpenMLPython},
Numpy \citep{numpy},
pandas \citep{pandas1}, \citep{pandas2},
PyTorch \citep{PyTorch},
scikit-learn \citep{scikit-learn},
SciPy \citep{SciPy},
pymoo \citep{pymoo}, and
tqdm \citep{tqdm}.
